---
title: "Mediation-analysis"
output:
  word_document: default
  html_document: default
  pdf_document: default
date: "2026-05-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Installation and loading script
install_and_load <- function() {
  
  # List of packages
  packages <- c(
    "lavaan",    # Structural equation modeling
    "dplyr",     # Data manipulation
    "tidyr",     # Data tidying
    "purrr",     # Functional programming
    "readr",     # Read CSV files
    "knitr",     # Pretty tables
    "openxlsx"   # Write Excel files
  )
  
  # Check which packages are missing
  missing <- packages[!packages %in% installed.packages()[,"Package"]]
  
  # Install missing packages
  if(length(missing) > 0) {
    cat("Installing missing packages:", paste(missing, collapse = ", "), "\n")
    install.packages(missing, dependencies = TRUE)
  } else {
    cat("All packages are already installed.\n")
  }
  
  # Load all packages
  cat("\nLoading packages...\n")
  for(pkg in packages) {
    library(pkg, character.only = TRUE)
    cat("✓ Loaded:", pkg, "\n")
  }
  
  cat("\n Setup complete! Ready for mediation analysis.\n")
}

# Run the setup
install_and_load()

```{r}
# Load required libraries
library(lavaan)
library(dplyr)
library(tidyr)
library(purrr)
library(readr)
library(knitr)
library(openxlsx)
```

```{r}
# Set path
commonality_path <- "/UK_BB/brainbody/commonality"

# Define modalities
brain_modalities <- c("allmri", "rs", "dwi", "smri")
body_modality <- "body"  # composite body

# Function to load and prepare data for mediation
load_mediation_data <- function(brain_mod, body_mod, data_path) {
  
  # File name pattern
  file_name <- paste0("g_obs_pred_brain_", brain_mod, "_body.csv")
  file_path <- file.path(data_path, file_name)
  
  if(!file.exists(file_path)) {
    warning(paste("File not found:", file_path))
    return(NULL)
  }
  
  data <- read_csv(file_path, show_col_types = FALSE)
  
  # Ddynamic renaming
  data <- data %>%
    rename(
      g_observed = g_obs_test,
      g_brain = all_of(paste0("g_pred_", brain_mod, "_test")),
      g_body = g_pred_body_test
    ) %>%
    
    # Z-score all three variables
    mutate(
      g_observed = scale(g_observed)[,1],
      g_brain    = scale(g_brain)[,1],
      g_body     = scale(g_body)[,1]
    )
  
  return(data)
}
```


```{r}
# Load all datasets
mediation_data_list <- list()

for(brain in brain_modalities) {
  cat("\nLoading data for brain modality:", brain, "\n")
  data <- load_mediation_data(brain, body_modality, commonality_path)
  
  if(!is.null(data)) {
    mediation_data_list[[brain]] <- data
    cat("  - Rows:", nrow(data), "\n")
    cat("  - Columns:", names(data), "\n")
  }
}
print(head(data), digits = 3)
```
                
```{r}
# Function to run mediation analysis
run_mediation <- function(data, brain_mod) {
  
  # Lavaan model specification
  # X = g_body (body prediction)
  # M = g_brain (brain prediction)
  # Y = g_observed (observed G factor)

  
  model <- '
    # Direct effect (c) and mediator effect (b)
    g_observed ~ b * g_brain + c * g_body
    
    # Path from body to brain (a)
    g_brain ~ a * g_body
    
    # Indirect effect (a * b)
    ab := a * b
    
    # Total effect (c + ab)
    total := c + ab
    
    # Proportion mediated
    prop := ab / total
  '
  
  # Fit model with bootstrap
  set.seed(1234)
  
  fit <- lavaan::sem(model, 
             data = data,
             test = "bollen.stine",
             se = "bootstrap", 
             bootstrap = 5000,
             #mimic = "MPlus"
             )  # Similar to MPlus defaults
  
  # Get parameter estimates with BCa confidence intervals
  params <- lavaan::parameterEstimates(fit, 
                               boot.ci.type = "bca.simple",
                               level = 0.95, 
                               ci = TRUE,
                               standardized = TRUE)
  
  kable(params)
            
  # Summary
  lav_sum = lavaan::summary(fit, 
                standardized = TRUE, 
                rsquare = TRUE, 
                fit.measures = TRUE)

  # Get fit measures
  fit_measures <- fitMeasures(fit, 
                              c("chisq", "df", "pvalue", "cfi", "rmsea", "srmr", "aic", "bic"))
  
  # Get R-squared values
  rsq <- inspect(fit, "r2")
  
  # Extract key results
  results <- list(
    brain_modality = brain_mod,
    sample_size = nrow(data),
    parameters = params,
    fit_measures = fit_measures,
    r_squared = rsq,
    lavaan_fit = fit,
    summary = lav_sum
  )
  
  return(results)
}
```

```{r}
# Run mediation
mediation_results <- lapply(
  names(mediation_data_list),
  function(brain_mod) run_mediation(mediation_data_list[[brain_mod]], brain_mod)
)

# Name the list elements
names(mediation_results) <- names(mediation_data_list)
```

Print parameter table for each modality
```{r}
for (mod in names(mediation_results)) {
  print(mediation_results[[mod]]$parameters)
}
```

Print R²
```{r}
for (mod in names(mediation_results)) {
  cat("\nR² for", mod, ":\n")
  print(mediation_results[[mod]]$r_squared)
}
```
Print fit measures
```{r}
for (mod in names(mediation_results)) {
  cat("\nFit measures for", mod, ":\n")
  print(mediation_results[[mod]]$fit_measures)
}
```

Combine all parameter estimates into one table (simple)
```{r}
library(dplyr)

all_params <- bind_rows(
  lapply(mediation_results, function(res) {
    res$parameters %>%
      mutate(brain_modality = res$brain_modality,
             sample_size = res$sample_size)
  })
)

print(all_params)
write.csv(all_params, "mediation_all_parameters.csv", row.names = FALSE)
```

age → (g_body, g_brain) → g_observed

```{r}
# Load combined file with age
combined_file <- file.path(commonality_path, "g_obs_pred_brain_plus_body_with_age.csv")

if (!file.exists(combined_file)) {
  stop("Combined file with age not found: g_obs_pred_brain_plus_body_with_age.csv")
}

df_all <- read_csv(combined_file, show_col_types = FALSE)

# Prepare data (rename + z-score)
data_age <- df_all %>%
  select(
    g_observed = g_obs,
    g_body     = g_pred_body,
    g_brain    = g_pred_brain,
    age        = Age
  ) %>%
  mutate(
    g_observed = scale(g_observed)[,1],
    g_body     = scale(g_body)[,1],
    g_brain    = scale(g_brain)[,1],
    age        = scale(age)[,1]
  )

# Mediation model
model_age <- '
  # outcome model
  g_observed  ~ b_g_body*g_body
  g_observed  ~ b_g_brain*g_brain
  g_observed  ~ c*age

  # mediator models
  g_body  ~ a_g_body*age
  g_brain ~ a_g_brain*age

  # indirect effects
  indirect_g_body  := a_g_body  * b_g_body
  indirect_g_brain := a_g_brain * b_g_brain
  sumIndirect      := indirect_g_body + indirect_g_brain

  # direct effect
  direct := c

  # total effect
  total := direct + sumIndirect

  # proportions
  prop_g_body      := indirect_g_body  / total
  prop_g_brain     := indirect_g_brain / total
  propSumIndirect  := sumIndirect      / total

  # allow mediators to covary
  g_body ~~ g_brain
'

# Fit model
set.seed(1234)
fit_age <- lavaan::sem(
  model_age,
  data = data_age,
  test = "bollen.stine",
  se = "bootstrap",
  bootstrap = 5000
)

# Extract parameters
params_age <- parameterEstimates(
  fit_age,
  boot.ci.type = "bca.simple",
  level = 0.95,
  ci = TRUE,
  standardized = TRUE
)

print(params_age)

# Save results
write.csv(params_age, "mediation_age_all_parameters.csv", row.names = FALSE)
```

